The \emph{Photochem} photochemical model

### 2022-07-06 ms.tex
finite volumes pretty much done

Consider an atmosphere that has been vertically divided into $m$ grid-cell (i.e. finite volumes) of thickness $\Delta z^j$. The superscript $j$ refers to a grid cell ($j$ does not refer to a power), where $j = 1$ is the lowest altitude grid-cell $j = m$ highest grid-cell. $z^j$ is the altitude at the center of grid-cell $j$, such that the upper and lower edges of the cell are given by $z^{j+1/2} = z^j + \frac{\Delta z^j}{2}$ and $z^{j-1/2} = z^j - \frac{\Delta z^j}{2}$, respectfully.

@@

We now must derive expressions for the gas and particle fluxes at grid-cell edges that are accurate and stable for our problem. For fluxes that are diffusion-dominated, a centered approximation is sufficient. However, for an advection-dominated flux, we can use a first-order upwind scheme, because a centered approximation is not stable for advection (Chapter 4 in \cite{Leveque_2002}).

\subsubsection{Gases}

@@

\begin{equation} \label{eq:phi_gas_m}
  \Phi_{i\text{,gas}}^{j-1/2} = - ( K + D_{i})^{j-1/2} n^{j-1/2} \frac{f_i^{j} - f_i^{j-1}}{\Delta z^{j-1/2}} - (\gamma_{i} n)^{j-1/2} \frac{\Delta z^{j-1}f_i^{j} + \Delta z^{j} f_i^{j-1}}{\Delta z^{j}+\Delta z^{j-1}}
\end{equation}
Substituting Equations \eqref{eq:phi_gas_p} and \eqref{eq:phi_gas_m} into Equation \eqref{eq:continuity_simple_approx} and re-arranging gives

\begin{equation}
\begin{split} 
  \frac{\partial f_{i \in \text{gas}}^j}{\partial t} &= (A_{i\text{,gas}}^{j\text{,upper}}+ B_{i\text{,gas}}^{j\text{,upper}}) f_{i}^{j+1} \\
  &+ (A_{i\text{,gas}}^{j\text{,center}} + B_{i\text{,gas}}^{j\text{,center}}) f_{i}^{j} \\
  &+ (A_{i\text{,gas}}^{j\text{,lower}} + B_{i\text{,gas}}^{j\text{,lower}}) f_{i}^{j-1} \\
  &+ \frac{\sigma_i^j}{n^{j}}
\end{split}
\end{equation}
Where, the $A$ coefficients (e.g. $A_{i\text{,gas}}^{j\text{,upper}}$) correspond to the diffusive approximations to the cell-edge fluxes:

\begin{align}
  A_{i\text{,gas}}^{j\text{,upper}} &= A_{i\text{,gas}}^{j,1} \nonumber \\
  A_{i\text{,gas}}^{j\text{,center}} &= - A_{i\text{,gas}}^{j,1} - A_{i\text{,gas}}^{j,2}  \nonumber \\
  A_{i\text{,gas}}^{j\text{,lower}} &= A_{i\text{,gas}}^{j,2} \nonumber
\end{align}
Where,

\begin{align}
  A_{i\text{,gas}}^{j,1} &= \frac{(K + D_{i})^{j+1/2} n^{j+1/2}}{n^j \Delta z^j \Delta z^{j+1/2}} \nonumber \\
  A_{i\text{,gas}}^{j,2} &= \frac{(K + D_{i})^{j-1/2} n^{j-1/2}}{n^j \Delta z^j \Delta z^{j-1/2}} \nonumber
\end{align}
Also, the $B$ coefficients (e.g. $B_{i\text{,gas}}^{j\text{,upper}}$) correspond to the advection approximations to the cell-edge fluxes:

\begin{align}
  B_{i\text{,gas}}^{j\text{,upper}} &= B_{i\text{,gas}}^{j,1} \nonumber \\
  B_{i\text{,gas}}^{j\text{,center}} &= B_{i\text{,gas}}^{j,2} + B_{i\text{,gas}}^{j,3}  \nonumber \\
  B_{i\text{,gas}}^{j\text{,lower}} &= B_{i\text{,gas}}^{j,4} \nonumber
\end{align}
Where,

\begin{align}
  B_{i\text{,gas}}^{j,1} &= \frac{(\gamma_{i} n)^{j+1/2}}{n^j \Delta z^j} \frac{\Delta z^{j}}{\Delta z^{j+1}+\Delta z^{j}} \nonumber \\
  B_{i\text{,gas}}^{j,2} &= \frac{(\gamma_{i} n)^{j+1/2}}{n^j \Delta z^j} \frac{\Delta z^{j+1}}{\Delta z^{j+1}+\Delta z^{j}} \nonumber \\
  B_{i\text{,gas}}^{j,3} &= - \frac{(\gamma_{i} n)^{j-1/2}}{n^j \Delta z^j} \frac{\Delta z^{j-1}}{\Delta z^{j}+\Delta z^{j-1}} \nonumber \\
  B_{i\text{,gas}}^{j,4} &= - \frac{(\gamma_{i} n)^{j-1/2}}{n^j \Delta z^j} \frac{\Delta z^{j}}{\Delta z^{j}+\Delta z^{j-1}} \nonumber 
\end{align}

The upper boundary condition is the flux through through the top edge of the highest-altitude cell, $\Phi_{i}^{m+1/2} = \Phi_{i}^\text{upper}$. Similarly, the lower boundary condition is the flux through the lower edge of the lowest-altitude cell, $\Phi_{i}^{1/2} = \Phi_{i}^\text{lower}$. It follows from Equation \eqref{eq:continuity_simple_approx} that

\begin{align}
  \frac{\partial f_i^m}{\partial t} &= - \frac{1}{n^{m}} \frac{\Phi_{i}^\text{upper} - \Phi_{i}^{m-1/2}}{\Delta z^m} + \frac{\sigma_i^m}{n^{m}} \\
  \frac{\partial f_i^1}{\partial t} &= - \frac{1}{n^{1}} \frac{\Phi_{i}^\text{1/2} - \Phi_{i}^\text{lower}}{\Delta z^1} + \frac{\sigma_i^1}{n^{1}}
\end{align}

\subsubsection{Particles}

>>>>

Consider an atmosphere that has been vertically divided into $m$ grid-cell (i.e. finite volumes) of thickness $\Delta z^j$. The superscript $j$ refers to a grid cell ($j$ does not refer to a power), where $j = l$ is the lowest altitude grid-cell $j = m$ highest grid-cell. $z^j$ is the altitude at the center of grid-cell $j$, such that the upper and lower edges of the cell are given by $z^{j+1/2} = z^j + \frac{\Delta z^j}{2}$ and $z^{j-1/2} = z^j - \frac{\Delta z^j}{2}$, respectfully.

@@

We now must derive expressions for the gas and particle fluxes at grid-cell edges that are accurate and stable for our problem.

% For fluxes that are diffusion-dominated, a centered approximation is sufficient. However, for an advection-dominated flux, we can use a first-order upwind scheme, because a centered approximation is not stable for advection (Chapter 4 in \cite{Leveque_2002}).

\subsubsection{Gases} \label{sec:FV_gases}

@@

\begin{equation} \label{eq:phi_gas_m}
  \Phi_{i\text{,gas}}^{j-1/2} = - ( K + D_{i})^{j-1/2} n^{j-1/2} \frac{f_i^{j} - f_i^{j-1}}{\Delta z^{j-1/2}} - (\gamma_{i} n)^{j-1/2} \frac{\Delta z^{j-1}f_i^{j} + \Delta z^{j} f_i^{j-1}}{\Delta z^{j}+\Delta z^{j-1}}
\end{equation}
Substituting Equations \eqref{eq:phi_gas_p} and \eqref{eq:phi_gas_m} into Equation \eqref{eq:continuity_simple_approx} and re-arranging gives an equation of the form

\begin{equation} \label{eq:dfdt_general}
\begin{split} 
  \frac{\partial f_{i}^j}{\partial t} &= (A_{i}^{j\text{,upper}}+ B_{i}^{j\text{,upper}}) f_{i}^{j+1} \\
  &+ (A_{i}^{j\text{,center}} + B_{i}^{j\text{,center}}) f_{i}^{j} \\
  &+ (A_{i}^{j\text{,lower}} + B_{i}^{j\text{,lower}}) f_{i}^{j-1} \\
  &+ \frac{\sigma_i^j}{n^{j}}
\end{split}
\end{equation}
% \begin{equation} \label{eq:dfdt_gas}
% \begin{split} 
%   \frac{\partial f_{i \in \text{gas}}^j}{\partial t} &= (A_{i\text{,gas}}^{j\text{,upper}}+ B_{i\text{,gas}}^{j\text{,upper}}) f_{i}^{j+1} \\
%   &+ (A_{i\text{,gas}}^{j\text{,center}} + B_{i\text{,gas}}^{j\text{,center}}) f_{i}^{j} \\
%   &+ (A_{i\text{,gas}}^{j\text{,lower}} + B_{i\text{,gas}}^{j\text{,lower}}) f_{i}^{j-1} \\
%   &+ \frac{\sigma_i^j}{n^{j}}
% \end{split}
% \end{equation}
The $A$ coefficients for gases correspond to the diffusive approximations to the cell-edge fluxes:

\begin{equation*}
\begin{aligned}
  A_{i\text{,gas}}^{j\text{,upper}} &= A_{i\text{,gas}}^{j,1} \\
  A_{i\text{,gas}}^{j\text{,center}} &= - A_{i\text{,gas}}^{j,1} - A_{i\text{,gas}}^{j,2} \\
  A_{i\text{,gas}}^{j\text{,lower}} &= A_{i\text{,gas}}^{j,2}
\end{aligned}
\quad\quad\quad
\begin{aligned}
  A_{i\text{,gas}}^{j,1} &= \frac{(K + D_{i})^{j+1/2} n^{j+1/2}}{n^j \Delta z^j \Delta z^{j+1/2}} \\
  A_{i\text{,gas}}^{j,2} &= \frac{(K + D_{i})^{j-1/2} n^{j-1/2}}{n^j \Delta z^j \Delta z^{j-1/2}} 
\end{aligned}
\end{equation*}
The $B$ coefficients for gases correspond to the advection approximations to the cell-edge fluxes:

\begin{equation*}
\begin{aligned}
  B_{i\text{,gas}}^{j\text{,upper}} &= B_{i\text{,gas}}^{j,1} \\
  B_{i\text{,gas}}^{j\text{,center}} &= B_{i\text{,gas}}^{j,2} + B_{i\text{,gas}}^{j,3} \\
  B_{i\text{,gas}}^{j\text{,lower}} &= B_{i\text{,gas}}^{j,4}
\end{aligned}
\quad\quad\quad
\begin{aligned}
  B_{i\text{,gas}}^{j,1} &= \frac{(\gamma_{i} n)^{j+1/2}}{n^j \Delta z^j} \frac{\Delta z^{j}}{\Delta z^{j+1}+\Delta z^{j}} \\
  B_{i\text{,gas}}^{j,2} &= \frac{(\gamma_{i} n)^{j+1/2}}{n^j \Delta z^j} \frac{\Delta z^{j+1}}{\Delta z^{j+1}+\Delta z^{j}} \\
  B_{i\text{,gas}}^{j,3} &= - \frac{(\gamma_{i} n)^{j-1/2}}{n^j \Delta z^j} \frac{\Delta z^{j-1}}{\Delta z^{j}+\Delta z^{j-1}} \\
  B_{i\text{,gas}}^{j,4} &= - \frac{(\gamma_{i} n)^{j-1/2}}{n^j \Delta z^j} \frac{\Delta z^{j}}{\Delta z^{j}+\Delta z^{j-1}} 
\end{aligned}
\end{equation*}
The lower boundary condition is the flux through the lower edge of the lowest-altitude cell, $\Phi_{i}^{l-1/2} = \Phi_{i}^\text{lower}$. Similarly, the upper boundary condition is the flux through through the top edge of the highest-altitude cell, $\Phi_{i}^{m+1/2} = \Phi_{i}^\text{upper}$. It follows from Equation \eqref{eq:continuity_simple_approx} that

\begin{align}
  \frac{\partial f_i^l}{\partial t} &= - \frac{1}{n^{l}} \frac{\Phi_{i}^\text{l+1/2} - \Phi_{i}^\text{lower}}{\Delta z^l} + \frac{\sigma_i^l}{n^{l}} \label{eq:lower_bc} \\
  \frac{\partial f_i^m}{\partial t} &= - \frac{1}{n^{m}} \frac{\Phi_{i}^\text{upper} - \Phi_{i}^{m-1/2}}{\Delta z^m} + \frac{\sigma_i^m}{n^{m}} \label{eq:upper_bc}
\end{align}
Plugging in Equation \eqref{eq:phi_gas_p} and \eqref{eq:phi_gas_m} into Equation \eqref{eq:lower_bc} and \eqref{eq:upper_bc} gives the following general expression for the rate mixing ratios change at the boundaries

\begin{equation} \label{eq:dfdt_boundary}
  \begin{aligned}
    \frac{\partial f_{i}^l}{\partial t} &= (A_{i}^{l\text{,upper}}+ B_{i}^{l\text{,upper}}) f_{i}^{l+1} \\
    &+ (A_{i}^{l\text{,center}} + B_{i}^{l\text{,center}}) f_{i}^{l} \\
    &+ \frac{\Phi_i^\text{lower}}{n^l \Delta z^l} \\
    &+ \frac{\sigma_i^l}{n^{l}}
  \end{aligned}
  \quad\quad\quad
  \begin{aligned}
    \frac{\partial f_{i}^m}{\partial t} &= (A_{i}^{m\text{,center}} + B_{i}^{m\text{,center}}) f_{i}^{m} \\
    &+ (A_{i}^{m\text{,lower}} + B_{i}^{m\text{,lower}}) f_{i}^{m-1} \\
    &- \frac{\Phi_i^\text{upper}}{n^m \Delta z^m} \\
    &+ \frac{\sigma_i^m}{n^{m}}
  \end{aligned}
  \end{equation}
% \begin{equation} \label{eq:dfdt_gas_boundary}
% \begin{aligned}
%   \frac{\partial f_{i \in \text{gas}}^l}{\partial t} &= (A_{i\text{,gas}}^{l\text{,upper}}+ B_{i\text{,gas}}^{l\text{,upper}}) f_{i}^{l+1} \\
%   &+ (A_{i\text{,gas}}^{l\text{,center}} + B_{i\text{,gas}}^{l\text{,center}}) f_{i}^{l} \\
%   &+ \frac{\Phi_i^\text{lower}}{n^l \Delta z^l} \\
%   &+ \frac{\sigma_i^l}{n^{l}}
% \end{aligned}
% \quad\quad\quad
% \begin{aligned}
%   \frac{\partial f_{i \in \text{gas}}^m}{\partial t} &= (A_{i\text{,gas}}^{m\text{,center}} + B_{i\text{,gas}}^{m\text{,center}}) f_{i}^{m} \\
%   &+ (A_{i\text{,gas}}^{m\text{,lower}} + B_{i\text{,gas}}^{m\text{,lower}}) f_{i}^{m-1} \\
%   &- \frac{\Phi_i^\text{upper}}{n^m \Delta z^m} \\
%   &+ \frac{\sigma_i^m}{n^{m}}
% \end{aligned}
% \end{equation}
For gases, the $A$ and $B$ coefficients for the lower boundary are

\begin{equation*}
  \begin{aligned}
  A_{i\text{,gas}}^{l\text{,upper}} &= A_{i\text{,gas}}^{l,1} \\
  A_{i\text{,gas}}^{l\text{,center}} &= - A_{i\text{,gas}}^{l,1} \\
  \end{aligned}
  \quad\quad
  \begin{aligned}
  B_{i\text{,gas}}^{l\text{,upper}} &= B_{i\text{,gas}}^{l,1} \\
  B_{i\text{,gas}}^{j\text{,center}} &= B_{i\text{,gas}}^{l,2} \\
  \end{aligned}
\end{equation*}
and the upper boundary,

\begin{equation*}
  \begin{aligned}
  A_{i\text{,gas}}^{m\text{,center}} &= - A_{i\text{,gas}}^{m,2} \\
  A_{i\text{,gas}}^{m\text{,lower}} &= A_{i\text{,gas}}^{m,2}
  \end{aligned}
  \quad\quad
  \begin{aligned}
  B_{i\text{,gas}}^{m\text{,center}} &= B_{i\text{,gas}}^{m,3} \\
  B_{i\text{,gas}}^{m\text{,lower}} &= B_{i\text{,gas}}^{m,4}
  \end{aligned}
\end{equation*}

\subsubsection{Particles} \label{sec:FV_particles}

The equation for particle flux (Equation \eqref{eq:phi_gas_1}), contains an advection term representing the falling of particles. This advection can be much stronger than the eddy diffusion term, especially high in the atmosphere. Therefore, for stability, we use a first-order upwind scheme for particle advection and a centered scheme for diffusion. A centered approximation is not stable for advection-dominated transport (Chapter 4 in \cite{Leveque_2002}). We can split the particle flux (Equation \eqref{eq:phi_particle_1}) into diffusive and advection components,

\begin{align*}
  \Phi_{i\text{,particle}} &= \Phi_{i\text{,diff}} + \Phi_{i\text{,advec}} \\
  \Phi_{i\text{,diff}} &= - Kn\frac{\partial f_{i}}{\partial z} \\
  \Phi_{i\text{,advec}} &= - w_i n f_{i}
\end{align*}
Therefore, we can write Equation \eqref{eq:continuity_simple_approx} as

\begin{equation}
  \frac{\partial f_{i\in\text{particle}}^j}{\partial t} = - \frac{1}{n^{j}} \frac{\Phi_{i\text{,diff}}^{j+1/2} - \Phi_{i\text{,diff}}^{j-1/2}}{\Delta z^j} - \frac{1}{n^{j}} \frac{\Phi_{i\text{,advec}}^{j+1/2} - \Phi_{i\text{,advec}}^{j-1/2}}{\Delta z^j} + \frac{\sigma_i^j}{n^{j}}
\end{equation}
Applying a centered scheme to the diffusive fluxes follows the same procedure as we did with gas diffusion (Section \ref{sec:FV_particles}). To apply a first-order upwind scheme, we first recognize that falling particles means information is traveling from up to down in the model. Therefore, an upwind scheme approximates advective fluxes with a forward difference

\begin{equation}
  \frac{\Phi_{i\text{,advec}}^{j+1/2} - \Phi_{i\text{,advec}}^{j-1/2}}{\Delta z^j} \approx \frac{\Phi_{i\text{,advec}}^{j+1} - \Phi_{i\text{,advec}}^{j}}{\Delta z^j} = \frac{(-w_i n f_i)^{j+1} - (-w_i n f_i)^{j}}{\Delta z^j}
\end{equation}
Applying a centered scheme to diffusion and an upwind scheme to advection, and rearranging, gives the $A$ and $B$ coefficients for Equation \eqref{eq:dfdt_general} for particles:

% \begin{equation} \label{eq:dfdt_particle}
% \begin{split} 
%   \frac{\partial f_{i \in \text{particle}}^j}{\partial t} &= (A_{i\text{,particle}}^{j\text{,upper}}+ B_{i\text{,particle}}^{j\text{,upper}}) f_{i}^{j+1} \\
%   &+ (A_{i\text{,particle}}^{j\text{,center}} + B_{i\text{,particle}}^{j\text{,center}}) f_{i}^{j} \\
%   &+ (A_{i\text{,particle}}^{j\text{,lower}} + B_{i\text{,particle}}^{j\text{,lower}}) f_{i}^{j-1} \\
%   &+ \frac{\sigma_i^j}{n^{j}}
% \end{split}
% \end{equation}
% Where,

\begin{equation*}
  \begin{aligned}
    A_{i\text{,particle}}^{j\text{,upper}} &= A_{i\text{,particle}}^{j,1} \\
    A_{i\text{,particle}}^{j\text{,center}} &= - A_{i\text{,particle}}^{j,1} - A_{i\text{,particle}}^{j,2} \\
    A_{i\text{,particle}}^{j\text{,lower}} &= A_{i\text{,particle}}^{j,2}
  \end{aligned}
  \quad\quad\quad
  \begin{aligned}
    A_{i\text{,particle}}^{j,1} &= \frac{K^{j+1/2} n^{j+1/2}}{n^j \Delta z^j \Delta z^{j+1/2}} \\
    A_{i\text{,particle}}^{j,2} &= \frac{K^{j-1/2} n^{j-1/2}}{n^j \Delta z^j \Delta z^{j-1/2}}
  \end{aligned}
\end{equation*}
% and,

\begin{equation*}
  \begin{aligned}
    B_{i\text{,particle}}^{j\text{,upper}} &= B_{i\text{,particle}}^{j,1} \\
    B_{i\text{,particle}}^{j\text{,center}} &= B_{i\text{,particle}}^{j,2} \\
    B_{i\text{,particle}}^{j\text{,lower}} &= 0
  \end{aligned}
  \quad\quad\quad
  \begin{aligned}
    B_{i\text{,particle}}^{j,1} &= \frac{w^{j+1}n^{j+1}}{n^j \Delta z^j} \\
    B_{i\text{,particle}}^{j,2} &= -\frac{ w^{j} }{\Delta z^j}
  \end{aligned}
\end{equation*}
Following the same procedure as we did in Section \ref{sec:FV_gases}, we can write the $A$ and $B$ coefficients for Equation \eqref{eq:dfdt_boundary} for particles at the lower boundary:

% \begin{equation} \label{eq:dfdt_particle_boundary}
% \begin{aligned}
%   \frac{\partial f_{i \in \text{particle}}^l}{\partial t} &= (A_{i\text{,particle}}^{l\text{,upper}}+ B_{i\text{,particle}}^{l\text{,upper}}) f_{i}^{l+1} \\
%   &+ (A_{i\text{,particle}}^{l\text{,center}} + B_{i\text{,particle}}^{l\text{,center}}) f_{i}^{l} \\
%   &+ \frac{\Phi_i^\text{lower}}{n^l \Delta z^l} \\
%   &+ \frac{\sigma_i^l}{n^{l}}
% \end{aligned}
% \quad\quad\quad
% \begin{aligned}
%   \frac{\partial f_{i \in \text{particle}}^m}{\partial t} &= (A_{i\text{,particle}}^{m\text{,center}} + B_{i\text{,particle}}^{m\text{,center}}) f_{i}^{m} \\
%   &+ (A_{i\text{,particle}}^{m\text{,lower}} + B_{i\text{,particle}}^{m\text{,lower}}) f_{i}^{m-1} \\
%   &- \frac{\Phi_i^\text{upper}}{n^m \Delta z^m} \\
%   &+ \frac{\sigma_i^m}{n^{m}}
% \end{aligned}
% \end{equation}
% The $A$ and $B$ coefficients for the lower boundary are

\begin{equation*}
  \begin{aligned}
  A_{i\text{,particle}}^{l\text{,upper}} &= A_{i\text{,particle}}^{l,1} \\
  A_{i\text{,particle}}^{l\text{,center}} &= - A_{i\text{,particle}}^{l,1} \\
  \end{aligned}
  \quad\quad
  \begin{aligned}
  B_{i\text{,particle}}^{l\text{,upper}} &= B_{i\text{,particle}}^{l,1} \\
  B_{i\text{,particle}}^{l\text{,center}} &= 0
  \end{aligned}
\end{equation*}
and the upper boundary,

\begin{equation*}
  \begin{aligned}
  A_{i\text{,particle}}^{m\text{,center}} &= - A_{i\text{,particle}}^{m,2} \\
  A_{i\text{,particle}}^{m\text{,lower}} &= A_{i\text{,particle}}^{m,2}
  \end{aligned}
  \quad\quad
  \begin{aligned}
  B_{i\text{,particle}}^{m\text{,center}} &= B_{i\text{,particle}}^{m,2} \\
  B_{i\text{,particle}}^{m\text{,lower}} &= 0
  \end{aligned}
\end{equation*}

\subsubsection{Summary}

Equations \eqref{eq:dfdt_general} and \eqref{eq:dfdt_boundary} are a system of ordinary differential equations (ODEs) which are a finite volume approximation of our original system of PDEs (Equation \eqref{eq:continuity_simple}). The number of ODEs is the number evolving species multiplied by the number of atmospheric grid-cells. We often use about 100 species and 200 vertical layers resulting in $\sim$20,000 ODEs.

\subsection{Solving the system of ordinary differential equations}

% At this point,

% Given initial conditions, boundary conditions, and values for various free parameters (e.g. $K$), we can evolve Equations \eqref{eq:dfdt_gas}, \eqref{eq:dfdt_gas_boundary}, \eqref{eq:dfdt_particle} and \eqref{eq:dfdt_particle_boundary} forward in time using a
